---
title: "Task 4 - Naples Reef Purple Urchins (Gracer)"
author: "Allison Gracer"
date: "10/26/2021"
output: 
  html_document: 
    toc: yes
    theme: spacelab
---

## Overview

**Data Citation:** Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(kableExtra)
```

## Setup

```{r}
# Attach packages
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(kableExtra)
```

## Read in the data

```{r}
urchins <- read_csv(here("data", "urchins.csv")) %>% # read in data using here()
  clean_names() # convert all column headers to lower snake case
```

## Wrangle the data

```{r}
purple_urchins_naples <- urchins %>% 
  mutate(date = mdy(date)) %>% # convert date class to date
  mutate(year = year(date)) %>%   # create a new column that only contains year
  filter(common_name == "Purple Urchin",
         year == "2010",
         site == "NAPL")
  
```

## Data visualization (exploratory)

### Histogram

```{r}
ggplot(data = purple_urchins_naples, 
       aes(x = size, fill = treatment)) + # create ggplot with size(cm) on x axis, differentiate fill based on treatment
  scale_fill_manual(values = c("paleturquoise3", "darkseagreen")) + # customize fill color
  scale_color_manual(values = c("paleturquoise4", "darkseagreen4")) + # customize outline color
  geom_histogram(bins = 12) + # specify ggplot type for histogram and customize bin size
  facet_wrap(~treatment) + # wrap data into 2 panels based on treatment
  labs( x = "Size (cm)") # customize graph labels
```


### QQ-plot

```{r}
ggplot(data = purple_urchins_naples, aes(sample = size)) + # create ggplot, specify data frame
  geom_qq() + # specify plot type as qq-plot
  facet_wrap(~treatment) # wrap data into 2 panels based on treatment
```

### Main takeaways

The histograms look like they may indicate a normal distribution for both treatment groups, but it is slightly unclear to me. However, after seeing the Q-Q plots both treatment groups appear to have a clear linear relationship. These exploratory graphs lead me to conclude that both treatment groups are normally distributed. 

## Descriptive statistics table

```{r}
purple_urchins_naples_stats <- purple_urchins_naples %>% # create a new data frame
  group_by(treatment) %>% # group data by treatment 
  summarize(
    mean = round(mean(size),2), # calculate mean, and round to 2 decimal places
    median = median(size), # calculate median
    sd = round(sd(size),2), # calculate standard deviation, and round to 2 decimal places
    sample_size = n()) # calculate sample size
```

## Finalized summary table

```{r}
kable(purple_urchins_naples_stats, # create simple table for rmarkdown with function kable()
      col.names = c("Treatment", "Mean(cm)", "Median(cm)", "Standard Deviation (cm)", "Sample Size(n)"), # specify column names
      align = "c", # center column alignment
      caption = "Table 1: Purple urchin size (cm) statistics for samples collected at Naples Reef in 2010") %>% 
  kable_paper() # changed table style
```

