---
title: "Task 3 - Santa Barbara Urchins (Gracer)"
author: "Allison Gracer"
date: "10/25/2021"
output: 
  html_document: 
    theme: spacelab
    toc: yes
---

## Overview

In this report Section 1 first compares sizes (cm) of purple and red sea urchins from data collected from 5 sites along the Santa Barbara Coast. Statistical analyses were run on this data and conclusions are discussed. Section 2 looks at the total number of observations of purple urchins by site by year. 

**Data Citation:** Reed, D. 2020. SBC LTER: Reef: Long-term experiment: Kelp removal: Urchin size frequency distribution ver 20. Environmental Data Initiative. https://doi.org/10.6073/pasta/5a1e9ef03aa47bd2225c0bb98a02a63b. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(janitor)
library(kableExtra)
library(lubridate)
```

## Setup

```{r}
# Attach packages
library(tidyverse)
library(here)
library(janitor)
library(kableExtra)
library(lubridate)
```

## Read in the data

```{r}
urchins <- read_csv(here("data", "urchins.csv")) %>% # read in data using here()
  clean_names() # convert all column headers to lower snake case
```

## Section 1

### Data Visualization: compare the distribution of sizes between purple and red sea urchins

###### To study trends in populations of sea urchins in the Santa Barbara Channel, we examine sample counts of urchins in each of five sites from 2008 to 2020.

```{r fig.cap = "Figure 1. Comparing size distribution of purple and red urchins from five sites in the Santa Barbara Channel based on measurements of urchin diameter (cm), measured to the nearest 0.5 cm."}
ggplot(data = urchins, aes(x = size)) + # create a ggplot using urchins data, with size(cm) on x axis
  geom_histogram(fill = "sienna3", # specify plot type as histogram, change graph colors, and bin size
                 color = "sienna4",
                 bins = 10) +
  facet_wrap(~ common_name) + # facet data based on common_name
  labs(x = "Urchin Diameter (cm)", # label x axis
       y = "Number of Urchins in Sample", # label y axis
       title = "Purple and Red Urchin Size(cm)") + # create a plot title
  theme(plot.title = element_text(hjust = 0.5)) # center plot title
  
```

### Descriptive statistics table: mean, median, standard deviation, and sample size of urchin size(cm)

###### Calculating mean, median, and standard deviation of urchin size(cm) to estimate the distribution of size between the two species. 


```{r}
urchins_stats <- urchins %>% # create a new data set called urchins_stats
  group_by(common_name) %>% # grouped by species (common_name)
  summarize(
    mean = round(mean(size),1), # calculate mean, and round to 2 decimal places
    median = median(size), # calculate median
    sd = round(sd(size),1), # calculate standard deviation, and round to 2 decimal places
    sample_size = n()) # calculate sample size
```

### Finalized summary table of statystical analysis (quantitative)

```{r}
kable(urchins_stats, # create simple table for rmarkdown with function kable()
      col.names = c("Common Name", "Mean(cm)", "Median(cm)", "Standard Deviation (cm)", "Sample Size(n)"), # specify column names
      align = "c", # center column alignment
      caption = "Table 1: Purple and red urchin size (cm) statistics for population samples collected along the Santa Barbara coast") %>% 
  kable_paper() # changed table style
```

### Overall takeaways from the statystical analysis (qualitative)

When comparing the two species, the data shows that overall red urchins are larger, with a mean size of 6.8 (cm) and median size of 7.0 (cm). The purple urchins are smaller with a mean size of 4.5(cm) and a median size of 4.5(cm).The difference in means between the two populations is 2.3(cm) and the percent difference is 66%. Additionally, it is important to note that the red urchins have a larger standard deviation 1.9 (cm) compared to the purple urchins 1.2 (cm). The standard deviation values allow us to conclude that the purple urchins have a higher central tendency (smaller SD value). This means the purple urchin sample has less variation from the mean size than the red urchin sample. 


## Section 2

### Data wrangling: exploring purple urchins by site and year

```{r}
purple_urchins <- urchins %>% # create a new data from called purple_urchins
  filter(common_name == "Purple Urchin") %>% #only include observations for Purple Urchins
  mutate(date = mdy(date)) %>% # convert date class to date
  mutate(year = year(date)) %>% # create a new column that only contains year
  mutate(site_full = case_when( #create a new column called site_full that contains full name for the 5 sites
    site %in% "NAPL" ~ "Naples Reef",
    site %in% "MOHK" ~ "Mohawk Reef",
    site %in% "CARP" ~ "Carpinteria Reef",
    site %in% "IVEE" ~ "Isla Vista Reef",
    site %in% "AQUE" ~ "Arroyo Quemado Reef"))
```

### Create a table: total number of purple urchins observed by site and year

```{r}
purple_urchins_site_year <- purple_urchins %>% # create new data frame to filter as needed for table
  group_by(site_full, year) %>%  # group data by site and year
  select(site_full, year) %>%  # select only site and year
  summarize(total_count = n())  # create new column with sum of counts per site per year
```

### Data visualization: counts of purple urchins by site and year


```{r fig.cap = "Figure 2. Total sample size (n) of purple urchins from five sites in the Santa Barbara Channel collected from 2008 to 2020."}
ggplot(data = purple_urchins_site_year, aes(x = year, y = total_count)) + # create a ggplot and assign axis
  geom_line(aes(color = site_full)) + # create a line graph, differentiating site names by color
  scale_x_continuous(breaks = 2008:2020) + # adjust the scale of my x-axis
  scale_y_continuous(breaks = c(250, 500, 750, 1000, 1250)) +
  scale_color_discrete(name = "Site") + # change legend title
  labs(x = "Year", # name x-axis label
     y = "Sample Size (n)", # name y-axis label
     title = "Purple Urchin Sample Size By Site Per Year", # add a plot title
     legend) +
  theme(plot.title = element_text(hjust = 0.5)) + # center plot title
  theme_light()
 
```


This graph may not be an accurate representation of changes in the purple urchin population. The data was collected from a combination of treatment areas. In the control group giant kelp species were left intact. In the annual group giant kelp was "removed...once each winter and allowed to re-colonize over the course of a year". This could lead to inaccurate representation of the overall population of purple urchins in ecosystems where there are not annual removals of giant kelp species. This graph combines all observations and does not take into account how many observations came from controlled vs. annual treatment sites. I would also have to know if there were issues such as time or funding restraints that resulted in fewer observations as the study progressed over time. The major drop off in samples taken from 2019-2020 could also be due to the Covid-19 pandemic forcing data collection to come to an abrupt stop.


